ANNO: The Anthropological Notation Ontology

### 2023-04-11 paper.tex
style: format intro source
improve intro

% For one author:
%\author{\inits{N.}\fnms{Name1} \snm{Surname1}\ead[label=e1]{first@somewhere.com}}
%\address{Department first, \orgname{University or Company name},
%Abbreviate US states, \cny{Country}\printead[presep={\\}]{e1}}
\todo{Autorenreihenfolge festlegen}
% Höffner, Heuschkel, Schmiedel, Penne, Fritzsch, Ludwig, Mohaupt, Labudde, Uciteli
% TODO: overleaf
% Two or more authors:
\begin{aug}
\author[B]{\inits{M.}\fnms{Marie} \snm{Heuschkel}\ead[label=e3]{x@somewhere.com}}
\author[A]{\inits{K.}\fnms{Konrad} \snm{Höffner}\ead[label=e1]{konrad.hoeffner@uni-leipzig.de}%
\thanks{Corresponding author. \printead{e1}.}}
\author[B]{\inits{F.}\fnms{Fabian} \snm{Schmiedel}\ead[label=e5]{a@somewhere.com}}
\author[B]{\inits{L.}\fnms{Laura} \snm{Penne}\ead[label=e8]{a@somewhere.com}}
\author[B]{\inits{H. T.}\fnms{Hamjo Tim} \snm{Fritzsch}\ead[label=e9]{c@somewhere.com}}
\author[B]{\inits{A.}\fnms{Andy} \snm{Ludwig}\ead[label=e2]{y@somewhere.com}}
\author[B]{\inits{M.}\fnms{Marleen} \snm{Mohaupt}\ead[label=e4]{z@somewhere.com}}
\author[B]{\inits{D.}\fnms{Dirk} \snm{Labudde}\ead[label=e6]{b@somewhere.com}}
\author[A]{\inits{A.}\fnms{Alexandr} \snm{Uciteli}\ead[label=e7]{c@somewhere.com}}
\address[A]{Institute for Medical Informatics, Statistics and Epidemiology (IMISE), \orgname{Leipzig University},
Saxony, \cny{Germany}\printead[presep={\\}]{e1,e7}}
\address[B]{\orgname{Hochschule Mittweida},
Saxony, \cny{Germany}\printead[presep={\\}]{e2,e3,e4,e5,e6}}
\end{aug}

@@

% KH: Ich hab hier mal den Text vom Poster als Basis für die Introduction genommen
\section{Introduction}\label{sec:introduction}
Historic and forensic anthropology gives insights into the behaviour, culture and way of life of humans of different epochs in the past.
Osteometry describes quantitative measurement on the human skeletton.
Anthropology relies on osteometric measurements, which has several problems:
Bones can only be investigated in place, which requires transport over far distances.
This causes wear in the bones, and can lead to different measurements.
This can be alleviated with a digital twin, which can be measured at any location.

@@

\section{Use Case: Integration into \aw{}}
% this is based on project report but has to be rewritten: less about AW3d and more about the usage of ANNO by it
ANNO is used by \aw{}~\citep{anthroworks3d}.
By combining user-friendly techniques of photogrammetry, insights from user experience research and knowledge from game development, a digital twin is created, which can subsequently be examined virtually.
This enables location-independent and parallel work without wear and tear on the bone.
The examination can be performed as often as desired, even if the skeletal individuals or collections are not available at the institute or have already been reburied.
It also draws its advantage when space is limited.
Thus, it requires only three SLR cameras, sufficient exposure and a computer.
The examination proceeds as follows: After photogrammetry is done and the 3D model is created, the objects are imported into the software and calibrated.
For calibration, there is an individual placeholder for each bone, which must first be selected.
Then the calibrated digital twin is placed in a placeholder skeleton.
Annotations can be made in the detailed view of each bone.
Point, line and area markings are available for selection.
Furthermore, it is possible to make measurements.
You can choose between distance, angle and circumference measurements.
After marking or measurement, an identification number is assigned to each one.
There is also a documentation about the time of the annotation as well as the name of the editor.
The marking or measurement can be classified according to certain categories (e.g. anatomical variation).
In addition, annotations are possible in a text field.
The annotations can be edited at any time afterwards, and the date of editing is recorded together with the author.
The annotations are subsequently displayed in different colors, so that overlaps can be kept apart.
A special feature of the software is the automation of measurements.
The program automatically sets pins for osteometric landmarks in a template view.
The person using the program can move these as desired.
The pins displayed have different color shades.
These differ depending on the relevance.
If the osteometric landmark is present in many measurement sections, its relevance increases and the pin appears darker.
After the pins have been roughly set, they can be refined using various views.
In the views, the individual osteometric landmarks are visible from different views, so that a fine adjustment can be made.
This is followed by the automatic measurement.
The individual values for each measurement section can be downloaded as a CSV file.
In addition, it is now possible to perform sex determinations using discriminant functions.
For this purpose, the required measurement sections can be selected.
The results are only visible in one display.
The automation allows a time saving, whereby more bones or skeletons can be examined in a shorter time.
Advantages over the previous approach are:

>>>>

% For one author:
%\author{\inits{N.}\fnms{Name1} \snm{Surname1}\ead[label=e1]{first@somewhere.com}}
%\address{Department first, \orgname{University or Company name},
%Abbreviate US states, \cny{Country}\printead[presep={\\}]{e1}}
\todo{Autorenreihenfolge festlegen}
% Höffner, Heuschkel, Schmiedel, Penne, Fritzsch, Ludwig, Mohaupt, Labudde, Uciteli
% TODO: overleaf
% Two or more authors:
\begin{aug}
\author[B]{\inits{M.}\fnms{Marie} \snm{Heuschkel}\ead[label=e3]{marie.heuschkel@hs-mittweida.de}}
\author[A]{\inits{K.}\fnms{Konrad} \snm{Höffner}\ead[label=e1]{konrad.hoeffner@uni-leipzig.de}%
\thanks{Corresponding author. \printead{e1}.}}
\author[B]{\inits{F.}\fnms{Fabian} \snm{Schmiedel}\ead[label=e5]{a@somewhere.com}}
\author[B]{\inits{L.}\fnms{Laura} \snm{Penne}\ead[label=e8]{a@somewhere.com}}
\author[B]{\inits{H. T.}\fnms{Hanjo Tim} \snm{Fritzsch}\ead[label=e9]{c@somewhere.com}}
\author[B]{\inits{A.}\fnms{Andy} \snm{Ludwig}\ead[label=e2]{y@somewhere.com}}% Fraunhofer AISEC ?
\author[B]{\inits{M.}\fnms{Marleen} \snm{Mohaupt}\ead[label=e4]{kreuzer@hs-mittweida.de}}
\author[B]{\inits{D.}\fnms{Dirk} \snm{Labudde}\ead[label=e6]{dirk.labudde@hs-mittweida.de }}
\author[A]{\inits{A.}\fnms{Alexandr} \snm{Uciteli}\ead[label=e7]{alexander.uciteli@imise.uni-leipzig.de}}
\address[A]{Institute for Medical Informatics, Statistics and Epidemiology (IMISE), \orgname{Leipzig University},
Saxony, \cny{Germany}\printead[presep={\\}]{e1,e7}}
\address[B]{\orgname{Hochschule Mittweida},
Saxony, \cny{Germany}\printead[presep={\\}]{e2,e3,e4,e5,e6}}
\end{aug}

@@

\section{Introduction}\label{sec:introduction}

Data from historical, prehistoric anthropology~\citep{prehistoricanthropology}, as well as modern forensic science, allow for a comprehensive understanding of deceased individuals from different time periods---
ranging from their identity and health status to aspects of their behaviour, lifestyle, culture and even concerning the circumstances of their death~\citep{spurensuche}.
For example, anthropological methods can be used to determine an individual's biological profile thus providing key information about their sex, age, height and ancestry and any further individualising traits~\citep{bioprofile}.

The science of anthropology can benefit from digitalisation in various way as digitisation processes facilitate location-independent and parallel work without the effects of wear and tear on the physical skeletal specimens, the material for investigation.
Currently problems of access could be resolved, as examinations can be conducted even if the skeletal material from individuals or collectives are not present at the institution or may have already been reburied.
The allocation of researchers and research material would greatly enhance the quality and extent of interdisciplinary collaborative research.
Moreover it allows for linking data of different study samples across geographical boundaries.
Data sets consolidated this way and adequately formalised would enable the application of efficient data analysing methods such as data mining and textmining techniques that could promote deeper insights.
Representing the anthropological domain through an ontology, such as the Anthropological Notation Ontology (ANNO) proposed here, is an appropriate formalisation~\citep{aw3dunpublished}.
Apart from contextual information, most informative clues can be found directly on the bone.
As a result, anthropological work relies on human anatomy, particularly the skeletal system and further tissues of the musculoskeletal system to classify and describe and thus document bones, anatomical structures and diagnostic features within the overall framework of the skeleton in a comprehensive and transparent fashion (Quelle 5).\todo{Heuschkel unveröffentlichte Forschung}

By transferring the principle of Earth topography, it is possible to create a precise map of the human skeletal and musculoskeletal system's \enquote{terrain} (QUELLE 6).\todo{z.B. Beleg zur Topographie. KH: könnte man hier als Quelle die TA nehmen?}
This map would contain visible anatomical surface structures as well as artificial objects such as measurement points or content-related or methodologically based classifications and boundaries.
These would be unequivocally located by accurate and exhaustive definitions that in turn would allow for their easy retrieval and moreover would serve as a basis for the objective examination, for instance in the form of measurements(QUELLE 7).\todo{Heuschkel unveröffentlichte Forschung}
Therefore the focus of the first research stage in the development of ANNO was therefore on the creation and usage of such a map.

@@

\section{Use Case: Integration into \aw{}}
% this is based on project report but has to be rewritten: less about AW3d and more about the usage of ANNO by it
ANNO is used by \aw{}~\citep{aw3d}.
By combining user-friendly techniques of photogrammetry, insights from user experience research and knowledge from game development, a digital twin is created, which can subsequently be examined virtually.
This enables location-independent and parallel work without wear and tear on the bone.
The examination can be performed as often as desired, even if the skeletal individuals or collections are not available at the institute or have already been reburied.
It also draws its advantage when space is limited.
Thus, it requires only three SLR cameras, sufficient exposure and a computer.
The examination proceeds as follows: After photogrammetry is done and the 3D model is created, the objects are imported into the software and calibrated.
For calibration, there is an individual placeholder for each bone, which must first be selected.
Then the calibrated digital twin is placed in a placeholder skeleton.
Annotations can be made in the detailed view of each bone.
Point, line and area markings are available for selection.
Furthermore, it is possible to make measurements.
You can choose between distance, angle and circumference measurements.
After marking or measurement, an identification number is assigned to each one.
There is also a documentation about the time of the annotation as well as the name of the editor.
The marking or measurement can be classified according to certain categories (e.g. anatomical variation).
In addition, annotations are possible in a text field.
The annotations can be edited at any time afterwards, and the date of editing is recorded together with the author.
The annotations are subsequently displayed in different colors, so that overlaps can be kept apart.
A special feature of the software is the automation of measurements.
The program automatically sets pins for osteometric landmarks in a template view.
The person using the program can move these as desired.
The pins displayed have different color shades.
These differ depending on the relevance.
If the osteometric landmark is present in many measurement sections, its relevance increases and the pin appears darker.
After the pins have been roughly set, they can be refined using various views.
In the views, the individual osteometric landmarks are visible from different views, so that a fine adjustment can be made.
This is followed by the automatic measurement.
The individual values for each measurement section can be downloaded as a CSV file.
In addition, it is now possible to perform sex determinations using discriminant functions.
For this purpose, the required measurement sections can be selected.
The results are only visible in one display.
The automation allows a time saving, whereby more bones or skeletons can be examined in a shorter time.
Advantages over the previous approach are:
